\vspace{-0.5in} \horrule{3pt} \\[0.4cm] \bf The competing benefits of noise and heterogeneity\\in neural coding \horrule{1pt}

### 2014-06-25 paper/manuscript.tex
Fixes to the figureless paper.
- Also some minor cosmetic edits to the references
Made a nicer preprint manuscript

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{times}
\usepackage{graphicx}
\usepackage{color}
\usepackage{multirow}

\usepackage[authoryear]{natbib}

@@

%%% margins
\textheight 23.4cm
\textwidth 14.65cm
\oddsidemargin 0.375in
\evensidemargin 0.375in
\topmargin  -0.55in

\renewcommand{\baselinestretch}{2}

\interfootnotelinepenalty=10000

% Different font in captions
\newcommand{\captionfonts}{\normalsize}

\makeatletter
\long\def\@makecaption#1#2{%
  \renewcommand{\baselinestretch}{1}% EH: stop captions running off the bottom of the page
  \vskip\abovecaptionskip
  \sbox\@tempboxa{{\captionfonts #1: #2}}%
  \ifdim \wd\@tempboxa >\hsize
    {\captionfonts #1: #2\par}
  \else
    \hbox to\hsize{\hfil\box\@tempboxa\hfil}%
  \fi
  \vskip\belowcaptionskip}
\makeatother
%%%%%

\usepackage{nameref}

@@

\begin{document}
\hspace{13.9cm}

\ \vspace{20mm}\\

{\LARGE The competing benefits of noise and heterogeneity in neural coding}

\ \\
{\bf \large Eric Hunsberger$^{1*}$, Matthew Scott$^2$, Chris Eliasmith$^1$}\\
{$^1$Centre for Theoretical Neuroscience, University of Waterloo, Waterloo, Canada.}\\
{$^2$Department of Applied Mathematics, University of Waterloo, Waterloo, Canada.}\\
{$^*$Corresponding author: ehunsber@uwaterloo.ca}\\
%

%\ \\[-2mm]
{\bf Keywords:} Heterogeneity; stochastic resonance; population coding

\thispagestyle{empty}
\markboth{}{NC instructions}
%
\ \vspace{-0mm}\\
%
%Abstract
\begin{center} {\bf Abstract} \end{center}

@@

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics{figure1_experiment.pdf}
  \fi
  \caption{
    \textbf{The experimental setup.} In our experiment, neurons are fed a common aperiodic input. Each neuron also has an independent white noise process and a randomly chosen bias that are added to the input. Varying the amplitude of the noise and the width of the interval from which the biases are randomly drawn allows us to control the amount of noise and heterogeneity in the experiment, respectively. Additionally, some ``off'' neurons invert their input (not shown). The output of each FHN or LIF neuron model is summed and filtered to produce the population output; this is compared with the input signal using mutual information as a metric.
  }
  \label{fig:exp}
\end{figure}

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure2_infohetero.pdf}
  \fi
  \caption{
    \textbf{Heterogeneity increases information content of a population of neurons, and exhibits a resonance effect.}
    Adding some heterogeneity (x-axis) to a population of LIF or FHN neurons increases their ability to represent an input signal (y-axis), while too much heterogeneity decreases representational ability. This effect is only observed at lower noise levels; at higher levels of noise, the noise alone is able to fully desynchronize neuronal firing and linearize neuron responses, so adding heterogeneity offers no additional benefit.
  }
  \label{fig:infohetero}
\end{figure}

@@

@@

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure3_infocontour.pdf}
  \fi
  \caption{
    \textbf{The benefits of noise and heterogeneity to information representation are not independent.}
We measured the information content (top, in bits) and information rate (bottom, in bits per spike),
across many levels of noise and heterogeneity.
In LIF neurons, both noise and heterogeneity show independent resonance effects:
given that only one of noise or heterogeneity is present,
there is a non-zero and finite level of noise or heterogeneity that optimizes information content.
Setting both noise and heterogeneity to their (independent) optimal levels
fails to further improve the representational ability of the population,
showing that these effects are not independent.
We propose that both parameters use
the same underlying mechanisms of desynchronization and linearization,
so once the population has been desynchronized and linearized by one parameter,
adding the other cannot further improve information content.
Heterogeneity is unable to fully desynchronize FHN neurons (see \textsc{\nameref{scn:desync}}),
so information content in these neurons is maximized for a specific combination of heterogeneity and noise. However, this effect is not seen when examining the information rate,
indicating that noise benefits neural systems partly through increasing firing rates.
  }
  \label{fig:infocontour}
\end{figure}

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure4_infonoise.pdf}
  \fi
  \caption{
    \textbf{Noise increases information content of homogeneous populations of FHN and LIF neurons.} A) In homogeneous populations, noise exhibits a resonance effect on information content. B) The resonance effect of noise is less pronounced when measuring the information rate, indicating that noise operates partially by increasing firing rates.
  }
  \label{fig:infonoise}
\end{figure}

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure5_phase.pdf}
  \fi
  \caption{
    \textbf{Noise and heterogeneity both desynchronize neurons.} To measure whether noise and heterogeneity have a desynchronizing effect on populations of neurons, we measured the standard deviation of neuron phases across the population (see \textsc{\nameref{scn:methods}}), where a smaller standard deviation indicates increased synchronization. Heterogeneity (individual traces) is seen to significantly desynchronize both FHN and LIF neurons, specifically across low to moderate levels of noise. At high levels of noise, the noise itself is sufficient to desynchronize neurons, and additional heterogeneity has no effect.
  }
  \label{fig:phase}
\end{figure}

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure6_syncraster.pdf}
  \fi
  \caption{
    \textbf{The effects of noise and heterogeneity on LIF neuron firing.}
    Spike rasters illustrate the effects of noise and heterogeneity parameters on neuronal firing in a population of 32 LIF neurons. With low noise and heterogeneity, neuronal firing is highly synchronized. Adding a moderate amount of noise or heterogeneity desynchronizes this firing. Adding a moderate level of both parameters cannot provide further desynchronization, since firing is already quite desynchronized. High levels of noise cause rapid firing in almost all neurons, degrading the ability of the population to represent the input signal. High levels of heterogeneity cause the signal to be entirely above or below the firing thresholds of many neurons, again degrading representational ability. Noise: $\sigma_\eta = $$10^{-3}$ (low), $10^{-2}$ (moderate), $10^{-1}$ (high). Heterogeneity: $b_r = $$10^{-2}$ (low), $10^{-1}$ (moderate), $10^{0}$ (high).
  }
  \label{fig:syncraster}
\end{figure}

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure7_tuningnoisy.pdf}
  \fi
  \caption{
    \textbf{Noise linearizes neuron tuning curves.}
    We measured the tuning (stimulus-response) curves of LIF and FHN neurons under various levels of noise.
    We observe that noise linearizes the neuron tuning curves,
    Specifically, noise eliminates the sharp threshold in the tuning curve by
    occasionally pushing the neuron above its firing threshold
    for stimuli that are below the threshold.
    Noise also increases neuron firing rates across some or all stimuli.
  }
  \label{fig:tuning}
\end{figure}

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure8_tuninghetero.pdf}
  \fi
  \caption{
    \textbf{Heterogeneity linearizes the response of a neuronal population to a stimulus.} A) Tuning curves for a heterogeneous population of $N = 9$ neurons. B) The population response of the heterogeneous population (heavy solid line) is much closer to a linear response (heavy dot-dashed line) than a single LIF tuning curve (heavy dashed line), even though the heterogeneous thresholds are far from evenly spaced.
  }
  \label{fig:tuninghetero}
\end{figure}

@@

\bibitem[Burton et~al., 2012]{Burton2012}
Burton, S.~D., Ermentrout, G.~B., and Urban, N.~N. (2012).
\newblock {Intrinsic heterogeneity in oscillatory dynamics limits
  correlation-induced neural synchronization.}
\newblock {\em Journal of neurophysiology}, 108(8):2115--33.

@@

\bibitem[Fitzhugh, 1961]{FitzHugh1961}
Fitzhugh, R. (1961).
\newblock {Impulses and physiological states in theoretical models of nerve
  membrane}.
\newblock {\em Biophysical journal}, 1(1948):445--466.

\bibitem[Gal\'{a}n et~al., 2008]{Galan2008}
Gal\'{a}n, R.~F., Ermentrout, G.~B., and Urban, N.~N. (2008).
\newblock {Optimal time scale for spike-time reliability: theory, simulations,
  and experiments.}
\newblock {\em Journal of neurophysiology}, 99(1):277--83.

@@

\bibitem[Hubel and Wiesel, 1962]{Hubel1962}
Hubel, D. and Wiesel, T. (1962).
\newblock {Receptive fields, binocular interaction and functional architecture
  in the cat's visual cortex}.
\newblock {\em The Journal of physiology}, 160:106--154.

@@

\bibitem[Marsat and Maler, 2010]{Marsat2010}
Marsat, G. and Maler, L. (2010).
\newblock {Neural heterogeneity and efficient population codes for
  communication signals.}
\newblock {\em Journal of neurophysiology}, 104(5):2543--55.

@@

\bibitem[Salinas and Abbott, 1994]{Salinas1994}
Salinas, E. and Abbott, L. (1994).
\newblock {Vector reconstruction from firing rates}.
\newblock {\em Journal of computational neuroscience}, 1(1):89--107.

@@

\bibitem[Stocks, 2000]{Stocks2000}
Stocks, N.~G. (2000).
\newblock {Suprathreshold stochastic resonance in multilevel threshold
  systems}.
\newblock {\em Physical review letters}, 84(11):2310--3.

@@

\bibitem[Tateno et~al., 2004]{Tateno2004}
Tateno, T., Harsch, a., and Robinson, H. P.~C. (2004).
\newblock {Threshold firing frequency-current relationships of neurons in rat
  somatosensory cortex: type 1 and type 2 dynamics.}
\newblock {\em Journal of neurophysiology}, 92(4):2283--94.

\bibitem[Tomita et~al., 1974]{Tomita1974}
Tomita, K., Ohta, T., and Tomita, H. (1974).
\newblock {Irreversible circulation and orbital revolution}.
\newblock {\em Prog. Theor. Phys}, 52(6):1744--1765.

\bibitem[Tsubo et~al., 2007]{Tsubo2007}
Tsubo, Y., Takada, M., Reyes, A.~D., and Fukai, T. (2007).
\newblock {Layer and frequency dependencies of phase response properties of
  pyramidal neurons in rat motor cortex.}
\newblock {\em The European journal of neuroscience}, 25(11):3429--41.

>>>>

\documentclass[letterpaper,11pt]{article}

\usepackage{amsmath}
\usepackage{times}
\usepackage{graphicx}
\usepackage{color}
\usepackage{multirow}

\usepackage[authoryear]{natbib}
\usepackage{nameref}

@@

\usepackage{url}
\renewcommand*{\UrlFont}{\ttfamily\small\relax}

\usepackage[top=1in, bottom=1in, left=1.5in, right=1.5in]{geometry}

@@

% Why does higher tolerance stop the overfull lines? But it does!
\tolerance=10000

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} 	% Horizontal rule

\title{
  \vspace{-0.5in}
  \horrule{3pt} \\[0.4cm]
  \bf The competing benefits of noise and heterogeneity\\in neural coding
  \horrule{1pt}
}

\newcommand{\makeauthor}[5]{
  \normalsize
  \begin{tabular}{c}
    \bf #1\\
    #2\\
    #3\\
    #4\\
    #5
  \end{tabular}
}

\author{
  \hspace{-1em}
  \makeauthor{Eric Hunsberger}
             {Computational Neuroscience Research Group}
             {University of Waterloo}
             {Waterloo, ON, Canada}
             {\url{ehunsber@uwaterloo.ca}}
  \makeauthor{Matthew Scott}
             {Department of Applied Mathematics}
             {University of Waterloo}
             {Waterloo, ON, Canada}
             {\relax} \\ \\
  \hspace{-1em}
  \makeauthor{Chris Eliasmith}
             {Computational Neuroscience Research Group}
             {University of Waterloo}
             {Waterloo, ON, Canada}
             {}
}

\date{\today}

\begin{document}

\maketitle

\begin{center} {\bf Abstract} \end{center}

@@

{\bf Keywords:} Heterogeneity; stochastic resonance; population coding

{\bf Source code:} \url{https://github.com/ctn-archive/hunsberger-neco2014}

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics{figure1_experiment.pdf}
  \else
    INSERT FIGURE 1 HERE
  \fi
  \caption{
    \textbf{The experimental setup.} In our experiment, neurons are fed a common aperiodic input. Each neuron also has an independent white noise process and a randomly chosen bias that are added to the input. Varying the amplitude of the noise and the width of the interval from which the biases are randomly drawn allows us to control the amount of noise and heterogeneity in the experiment, respectively. Additionally, some ``off'' neurons invert their input (not shown). The output of each FHN or LIF neuron model is summed and filtered to produce the population output; this is compared with the input signal using mutual information as a metric.
  }
  \label{fig:exp}
\end{figure}

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure2_infohetero.pdf}
  \else
    INSERT FIGURE 2 HERE
  \fi
  \caption{
    \textbf{Heterogeneity increases information content of a population of neurons, and exhibits a resonance effect.}
    Adding some heterogeneity (x-axis) to a population of LIF or FHN neurons increases their ability to represent an input signal (y-axis), while too much heterogeneity decreases representational ability. This effect is only observed at lower noise levels; at higher levels of noise, the noise alone is able to fully desynchronize neuronal firing and linearize neuron responses, so adding heterogeneity offers no additional benefit.
  }
  \label{fig:infohetero}
\end{figure}

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure3_infocontour.pdf}
  \else
    INSERT FIGURE 3 HERE
  \fi
  \caption{
    \textbf{The benefits of noise and heterogeneity to information representation are not independent.}
We measured the information content (top, in bits) and information rate (bottom, in bits per spike),
across many levels of noise and heterogeneity.
In LIF neurons, both noise and heterogeneity show independent resonance effects:
given that only one of noise or heterogeneity is present,
there is a non-zero and finite level of noise or heterogeneity that optimizes information content.
Setting both noise and heterogeneity to their (independent) optimal levels
fails to further improve the representational ability of the population,
showing that these effects are not independent.
We propose that both parameters use
the same underlying mechanisms of desynchronization and linearization,
so once the population has been desynchronized and linearized by one parameter,
adding the other cannot further improve information content.
Heterogeneity is unable to fully desynchronize FHN neurons (see \textsc{\nameref{scn:desync}}),
so information content in these neurons is maximized for a specific combination of heterogeneity and noise. However, this effect is not seen when examining the information rate,
indicating that noise benefits neural systems partly through increasing firing rates.
  }
  \label{fig:infocontour}
\end{figure}

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure4_infonoise.pdf}
  \else
    INSERT FIGURE 4 HERE
  \fi
  \caption{
    \textbf{Noise increases information content of homogeneous populations of FHN and LIF neurons.} A) In homogeneous populations, noise exhibits a resonance effect on information content. B) The resonance effect of noise is less pronounced when measuring the information rate, indicating that noise operates partially by increasing firing rates.
  }
  \label{fig:infonoise}
\end{figure}

@@

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure5_phase.pdf}
  \else
    INSERT FIGURE 5 HERE
  \fi
  \caption{
    \textbf{Noise and heterogeneity both desynchronize neurons.} To measure whether noise and heterogeneity have a desynchronizing effect on populations of neurons, we measured the standard deviation of neuron phases across the population (see \textsc{\nameref{scn:methods}}), where a smaller standard deviation indicates increased synchronization. Heterogeneity (individual traces) is seen to significantly desynchronize both FHN and LIF neurons, specifically across low to moderate levels of noise. At high levels of noise, the noise itself is sufficient to desynchronize neurons, and additional heterogeneity has no effect.
  }
  \label{fig:phase}
\end{figure}

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure6_syncraster.pdf}
  \else
    INSERT FIGURE 6 HERE
  \fi
  \caption{
    \textbf{The effects of noise and heterogeneity on LIF neuron firing.}
    Spike rasters illustrate the effects of noise and heterogeneity parameters on neuronal firing in a population of 32 LIF neurons. With low noise and heterogeneity, neuronal firing is highly synchronized. Adding a moderate amount of noise or heterogeneity desynchronizes this firing. Adding a moderate level of both parameters cannot provide further desynchronization, since firing is already quite desynchronized. High levels of noise cause rapid firing in almost all neurons, degrading the ability of the population to represent the input signal. High levels of heterogeneity cause the signal to be entirely above or below the firing thresholds of many neurons, again degrading representational ability. Noise: $\sigma_\eta = $$10^{-3}$ (low), $10^{-2}$ (moderate), $10^{-1}$ (high). Heterogeneity: $b_r = $$10^{-2}$ (low), $10^{-1}$ (moderate), $10^{0}$ (high).
  }
  \label{fig:syncraster}
\end{figure}

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure7_tuningnoisy.pdf}
  \else
    INSERT FIGURE 7 HERE
  \fi
  \caption{
    \textbf{Noise linearizes neuron tuning curves.}
    We measured the tuning (stimulus-response) curves of LIF and FHN neurons under various levels of noise.
    We observe that noise linearizes the neuron tuning curves,
    Specifically, noise eliminates the sharp threshold in the tuning curve by
    occasionally pushing the neuron above its firing threshold
    for stimuli that are below the threshold.
    Noise also increases neuron firing rates across some or all stimuli.
  }
  \label{fig:tuning}
\end{figure}

@@

\begin{figure}
  \ifx\hidefigures\undefined
    \centering
    \includegraphics[width=\textwidth]{figure8_tuninghetero.pdf}
  \else
    INSERT FIGURE 8 HERE
  \fi
  \caption{
    \textbf{Heterogeneity linearizes the response of a neuronal population to a stimulus.} A) Tuning curves for a heterogeneous population of $N = 9$ neurons. B) The population response of the heterogeneous population (heavy solid line) is much closer to a linear response (heavy dot-dashed line) than a single LIF tuning curve (heavy dashed line), even though the heterogeneous thresholds are far from evenly spaced.
  }
  \label{fig:tuninghetero}
\end{figure}

@@

\bibitem[Burton et~al., 2012]{Burton2012}
Burton, S.~D., Ermentrout, G.~B., and Urban, N.~N. (2012).
\newblock {Intrinsic heterogeneity in oscillatory dynamics limits
  correlation-induced neural synchronization.}
\newblock {\em Journal of Neurophysiology}, 108(8):2115--33.

@@

\bibitem[Fitzhugh, 1961]{FitzHugh1961}
Fitzhugh, R. (1961).
\newblock {Impulses and physiological states in theoretical models of nerve
  membrane}.
\newblock {\em Biophysical Journal}, 1(1948):445--466.

\bibitem[Gal\'{a}n et~al., 2008]{Galan2008}
Gal\'{a}n, R.~F., Ermentrout, G.~B., and Urban, N.~N. (2008).
\newblock {Optimal time scale for spike-time reliability: theory, simulations,
  and experiments.}
\newblock {\em Journal of Neurophysiology}, 99(1):277--83.

@@

\bibitem[Hubel and Wiesel, 1962]{Hubel1962}
Hubel, D. and Wiesel, T. (1962).
\newblock {Receptive fields, binocular interaction and functional architecture
  in the cat's visual cortex}.
\newblock {\em The Journal of Physiology}, 160:106--154.

@@

\bibitem[Marsat and Maler, 2010]{Marsat2010}
Marsat, G. and Maler, L. (2010).
\newblock {Neural heterogeneity and efficient population codes for
  communication signals.}
\newblock {\em Journal of Neurophysiology}, 104(5):2543--55.

@@

\bibitem[Salinas and Abbott, 1994]{Salinas1994}
Salinas, E. and Abbott, L. (1994).
\newblock {Vector reconstruction from firing rates}.
\newblock {\em Journal of Computational Neuroscience}, 1(1):89--107.

@@

\bibitem[Stocks, 2000]{Stocks2000}
Stocks, N.~G. (2000).
\newblock {Suprathreshold stochastic resonance in multilevel threshold
  systems}.
\newblock {\em Physical Review Letters}, 84(11):2310--3.

@@

\bibitem[Tateno et~al., 2004]{Tateno2004}
Tateno, T., Harsch, a., and Robinson, H. P.~C. (2004).
\newblock {Threshold firing frequency-current relationships of neurons in rat
  somatosensory cortex: type 1 and type 2 dynamics.}
\newblock {\em Journal of Neurophysiology}, 92(4):2283--94.

\bibitem[Tomita et~al., 1974]{Tomita1974}
Tomita, K., Ohta, T., and Tomita, H. (1974).
\newblock {Irreversible circulation and orbital revolution}.
\newblock {\em Progress of Theoretical Physics}, 52(6):1744--1765.

\bibitem[Tsubo et~al., 2007]{Tsubo2007}
Tsubo, Y., Takada, M., Reyes, A.~D., and Fukai, T. (2007).
\newblock {Layer and frequency dependencies of phase response properties of
  pyramidal neurons in rat motor cortex.}
\newblock {\em European Journal of Neuroscience}, 25(11):3429--41.
